---
author: "Sarah Salter"
date: "October 22nd, 2016"
output: html_document
---

<center> <h1> Fatal Laws: </h1> </center>
<center> <h1> *Present Day Safety Laws & Their Association to Traffic Fatalities in the United States* </h1> </center>
<style>p { color: black; font-size: 16px }
.paragraph { font-size: 20px; }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##I. Introduction 

Yearly traffic related accidents currently contribute to the 4th leading cause of death in the United States, and the leading cause of death for persons between the ages of 1 to 44 [1]. The National Highway Traffic Safety Administration (NHTSA) estimates that highway crashes alone have an annual price tag of around $871 billion in economic loss and social harm [3].  According to the NHTSA, speeding, drunken driving, and distracted driving are the three main behavioral factors contributing to motor vehicle related fatalities across the nation.  It was found that these three behaviors alone comprise approximately 56% of the economic loss and 62% of the societal harm related to annual highway crashes, in addition to the average of 30,000+ lives lost in recent years. 

Due to the fact that traffic related fatalities are preventable deaths directly related to human error, state legislators have passed numerous traffic safety regulations and launched many initiatives in an attempt to correct for this human error.  Since the establishment of the US Department of Transportation in 1967 and NHTSA in 1970, many federal/state laws and ad-campaigns have been established to "save lives, prevent injuries, and reduce vehicle-related crashes", as described by their mission statement.  While there have been great reductions in the amount of annual highway crashes over the past 4 decades, the downward trend of motor vehicle related death has shown to lack uniformity amongst the states [5]. As a result, this has prompted the questions: What effect do traffic safety laws have on the number of yearly accidents resulting in death within the US? In other words, can a law's non-existence or lax nature be indirectly fatal, for persons of that residing state, by 'allowing' for greater amounts of human error?

This paper examines the associations related to current state laws and traffic fatalities within the US in the year 2015. By controlling for several state-specific factors contributing to motor-vehicles accidents, I aim to explore the similarities and differences in fatality counts directly related to state traffic legislation.  Specifically, I attempt to explore whether the non-uniformity in state laws is directly related to the difference in fatality reduction? And if so, what state laws appear to have the greatest contribution to this difference?  And if not, what other contributing factors are responsible for the lack consistency state-wide? I hypothesize that similar states (based on non-traffic related characteristics) with more stringent laws and greater law enforcement will have a smaller amount of motor vehicle deaths then states that do not.  

##II. Data
###A. Fatality Analysis Reporting System Data
The Fatality Analysis Reporting System (FARS) data, created by the National Highway Traffic Safety Administration (NHTSA), can be defined as a nationwide census of fatal traffic crashes observed within the United States, as well as the District of Columbia and Puerto Rico.  According to NHTSA, for a crash to be eligible for this database the crash had to involve “at least one motor vehicle traveling on a traffic way customarily open to the public, and must have resulted in the death of a motorist or a non-motorist within 30 days of the crash” [6]. Fatalities are defined this way to account for crashes that may not have resulted in instantaneous death. Thus, it is assumed that people who are involved in a serious motor-vehicle accident and die up to 31 days after their accident die from accident related trauma/complications. Additionally, motor vehicles are defined as any motor powered vehicle, e.g.- automobile, truck, bus, motorcycle, bicycle, snowmobile, all-terrain vehicle, moped, golf cart, fork-lift, construction equipment, etc. 

FARS has been operational since 1975, and contains 20+ databases consisting of information in the form of accident-level, person-level, vehicle-level, and event-level data.  For the purposes of my research I chose to evaluate information only pertaining to 50 states and the District of Columbia. In my exploratory analysis, I specifically evaluated accident level data between 1975-2015. For further exploratory analysis, using all data levels I chose to only evaluate the years of 2010-2015. The reason for this is because the downward trend seemed pretty consistent in the years prior, and also because the variables evaluated in in 2010 and later were more consistent than the variables evaluate prior to 2010.  Thus, this was the best way to evaluate the greatest amount of similar variables across time, without needing to do techniques such as imputation for missing variable data. Furthermore, in my main the analysis, I chose to only use the 2015 accident-level data which allowed me to identify the number of accidents occurring in each state and the number of fatalities resulting from those accidents.  

###B. State Laws

Due to the fact that my research question involves state laws pertaining to traffic safety, I used many resources to collect laws for each state that I believed were established for traffic safety purposes and could potentially minimize person risk if obeyed. I identified the main areas of laws I was interested in such as: cellphone, seatbelt, child-restraint, speeding, DUI, young driver, older driver, helmet, red-light running, and aggressive driving. I then selected which laws I wanted to represent these areas of interest, e.g.- primary enforced laws related to cell phones, primary forced laws related to seatbelts, minimum age for learner's permit holders, maximum speed limit on urban/rural roads, frequency of proof of vision checks for older adults, etc. I identified 59 laws in total. Afterwards, I researched how each law was currently being implemented (in 2015) for all 50 state, and then I coded each state-specific implementation.  One of the main sources I used to obtain this information was the Highway Loss Data Institute from the Insurance Institute for Highway Safety (IIHS).  Here I obtained the information for 47 of the 59 total variables.

Specifically, all laws, besides those related to seatbelt enforcement, speeding, DUI Blood Alcohol Content for people under 21 years of age, DUI checkpoints, legalization of marijuana, and aggressive driving were obtained from IIHS.  Thus, a total of 12 out of the 59 variables were obtained from other sources.  (See Data Section of the Supplemental Page for more information).

###C. Motor Vehicle Accident Fatality Control

Next, I identified various motor vehicle accident risk factors that cannot be influenced directly by traffic legislation.  The purpose of identifying these variable is to determine what variables should be controlled for to ensure the best chance in evaluating traffic fatality associations related to legislation with minimum confounding variables not accounted for. 

For instance, first I assumed that traffic fatalities may differ in areas where there is higher unemployment.  For example, maybe those without jobs or frequently laid off from jobs display 'riskier' behave, e.g.- drive at faster speeds, drink and drive, do not wear their seatbelts, more often than people that are employed. As a result, I evaluate the unemployment rate for each state at each month in 2015, and calculated each state's median annual unemployment to account for this potential confounder.  Secondly, I considered the possibility that areas with higher income may have a greater amount of younger drivers with their own vehicles, and therefore a larger amount of inexperienced drivers on the road.  Since young drivers contribute a large amount to total accident fatalities, especially young drivers under the influence, I decided to control for this variable using the (Income Per Capita/1000) for each state.
Third, I assumed that states that have a greater amount of vehicle miles traveled will have a greater risk at accident related fatalities than states that whose population drives less.  Thus, I obtained total vehicle miles traveled for 2015 and divided this quantity by 1000 to make it more interpretable.  

Lastly, I assumed that seasonality would greatly contribute to a state's amount of traffic fatalities. For instance, I assume that states that experience greater amounts of snow and rain will most likely have a greater amount of risk than states that rarely experience any type of inclement weather.  I attempted to evaluate seasonality by using a yearly average temperature variable and a low/high snow variable which identifies states that have low-high amount of snow compared to little or no yearly snow fall.  

###D. Population Quantities

Population Projections from the Census/CDC

Population Growth

Population Percentage of 'Youth' (14-24) 

Population Percentage of 'Old' >65

```{r, echo=FALSE}
#*****************************************************
#1. Read in Accident Level Data Sets from 2010-2015
#2. Compare Colnames Across All Datasets to Determine which Variables are Consistent Acorss All 5 Years.
#3. Subset Datasets from Each Year to Include Only the Consistent Variables Evaluated at Every year Between 2010-2015.
#4. Rowbind Datasets On Like Variables (Arranged in the Same Order) to Create One Dataeset from Years 2010-2015.
#5. Plot Exploratory Plot to See the Patterns in Accident Level Data across Time.
#****************************************************

#Load necessary packages
library(foreign)
library(haven)

#Unzip & Download Data File 
#Read Accident Dataset from Years 2010-2015 
accident_data2015 = read_sas("~/Downloads/FARS_Accident2015.sas7bdat")
accident_data2014 = read_sas("~/Downloads/FARS_Accident2014.sas7bdat")
accident_data2013 = read_sas("~/Downloads/FARS_Accident2013.sas7bdat")
accident_data2012 = read_sas("~/Downloads/FARS_Accident2012.sas7bdat")
accident_data2011 = read_sas("~/Downloads/FARS_Accident2011.sas7bdat")
accident_data2010 = read_sas("~/Downloads/FARS_Accident2010.sas7bdat")

#*****************************************************
#Compare all column names 
#*****************************************************

accident_colnames_15 <- colnames(accident_data2015)
#length(accident_colnames_15) #52
#print(accident_colnames_15)
# [1] "STATE"     "ST_CASE"     "VE_TOTAL"   "VE_FORMS"    "PVH_INVL"   "PEDS"       "PERNOTMVIT"  "PERMVIT"    "PERSONS"    "COUNTY"    
# [11] "CITY"       "DAY"        "MONTH"      "YEAR"       "DAY_WEEK"   "HOUR"       "MINUTE"      "NHS"        "RUR_URB"    "FUNC_SYS"  
# [21] "RD_OWNER"   "ROUTE"      "TWAY_ID"    "TWAY_ID2"   "MILEPT"     "LATITUDE"   "LONGITUD"    "SP_JUR"     "HARM_EV"    "MAN_COLL"  
# [31] "RELJCT1"    "RELJCT2"    "TYP_INT"    "WRK_ZONE"   "REL_ROAD"   "LGT_COND"   "WEATHER1"   "WEATHER2"   "WEATHER"    "SCH_BUS"   
# [41] "RAIL"       "NOT_HOUR"   "NOT_MIN"    "ARR_HOUR"   "ARR_MIN"    "HOSP_HR"    "HOSP_MN"    "CF1"        "CF2"        "CF3"       
# [51] "FATALS"     "DRUNK_DR"

accident_colnames_14 <- colnames(accident_data2014)
#length(accident_colnames_14) #50
#print(accident_colnames_14)
# [1] "STATE"      "ST_CASE"    "VE_TOTAL"   "VE_FORMS"   "PVH_INVL"    "PEDS"       "PERNOTMVIT" "PERMVIT"    "PERSONS"    "COUNTY"    
# [11] "CITY"       "DAY"        "MONTH"      "YEAR"       "DAY_WEEK"   "HOUR"       "MINUTE"     "NHS"        "ROAD_FNC"   "ROUTE"     
# [21] "TWAY_ID"    "TWAY_ID2"   "MILEPT"     "LATITUDE"   "LONGITUD"   "SP_JUR"     "HARM_EV"    "MAN_COLL"   "RELJCT1"    "RELJCT2"   
# [31] "TYP_INT"    "WRK_ZONE"   "REL_ROAD"   "LGT_COND"   "WEATHER1"   "WEATHER2"   "WEATHER"    "SCH_BUS"    "RAIL"       "NOT_HOUR"  
# [41] "NOT_MIN"    "ARR_HOUR"   "ARR_MIN"    "HOSP_HR"    "HOSP_MN"    "CF1"        "CF2"        "CF3"        "FATALS"     "DRUNK_DR"  

accident_colnames_13 <- colnames(accident_data2013)
#length(accident_colnames_13) #50
#print(accident_colnames_13)
# [1] "STATE"      "ST_CASE"    "VE_TOTAL"   "VE_FORMS"   "PVH_INVL"    "PEDS"       "PERNOTMVIT" "PERMVIT"    "PERSONS"    "COUNTY"    
# [11] "CITY"       "DAY"        "MONTH"      "YEAR"       "DAY_WEEK"   "HOUR"       "MINUTE"     "NHS"        "ROAD_FNC"   "ROUTE"     
# [21] "TWAY_ID"    "TWAY_ID2"   "MILEPT"     "LATITUDE"   "LONGITUD"   "SP_JUR"     "HARM_EV"    "MAN_COLL"   "RELJCT1"    "RELJCT2"   
# [31] "TYP_INT"    "WRK_ZONE"   "REL_ROAD"   "LGT_COND"   "WEATHER1"   "WEATHER2"   "WEATHER"    "SCH_BUS"    "RAIL"       "NOT_HOUR"  
# [41] "NOT_MIN"    "ARR_HOUR"   "ARR_MIN"    "HOSP_HR"    "HOSP_MN"    "CF1"        "CF2"        "CF3"        "FATALS"     "DRUNK_DR"  

accident_colnames_12 <- colnames(accident_data2012)
#length(accident_colnames_12) #50
#print(accident_colnames_12)
# [1] "STATE"      "ST_CASE"    "VE_TOTAL"   "VE_FORMS"   "PVH_INVL"    "PEDS"       "PERNOTMVIT" "PERMVIT"    "PERSONS"    "COUNTY"    
# [11] "CITY"       "DAY"        "MONTH"      "YEAR"       "DAY_WEEK"   "HOUR"       "MINUTE"     "NHS"        "ROAD_FNC"   "ROUTE"     
# [21] "TWAY_ID"    "TWAY_ID2"   "MILEPT"     "LATITUDE"   "LONGITUD"   "SP_JUR"     "HARM_EV"    "MAN_COLL"   "RELJCT1"    "RELJCT2"   
# [31] "TYP_INT"    "WRK_ZONE"   "REL_ROAD"   "LGT_COND"   "WEATHER1"   "WEATHER2"   "WEATHER"    "SCH_BUS"    "RAIL"       "NOT_HOUR"  
# [41] "NOT_MIN"    "ARR_HOUR"   "ARR_MIN"    "HOSP_HR"    "HOSP_MN"    "CF1"        "CF2"        "CF3"        "FATALS"     "DRUNK_DR"  

accident_colnames_11 <- colnames(accident_data2011)
#length(accident_colnames_11) #50
#print(accident_colnames_11)
# [1] "STATE"      "ST_CASE"    "VE_TOTAL"   "VE_FORMS"   "PVH_INVL"    "PEDS"       "PERNOTMVIT" "PERMVIT"    "PERSONS"    "COUNTY"    
# [11] "CITY"       "DAY"        "MONTH"      "YEAR"       "DAY_WEEK"   "HOUR"       "MINUTE"     "NHS"        "ROAD_FNC"   "ROUTE"     
# [21] "TWAY_ID"    "TWAY_ID2"   "MILEPT"     "LATITUDE"   "LONGITUD"   "SP_JUR"     "HARM_EV"    "MAN_COLL"   "RELJCT1"    "RELJCT2"   
# [31] "TYP_INT"    "WRK_ZONE"   "REL_ROAD"   "LGT_COND"   "WEATHER1"   "WEATHER2"   "WEATHER"    "SCH_BUS"    "RAIL"       "NOT_HOUR"  
# [41] "NOT_MIN"    "ARR_HOUR"   "ARR_MIN"    "HOSP_HR"    "HOSP_MN"    "CF1"        "CF2"        "CF3"        "FATALS"     "DRUNK_DR"  


accident_colnames_10 <- colnames(accident_data2010)
#length(accident_colnames_10) #47
#print(accident_colnames_10)
# [1]  "STATE"    "ST_CASE"  "VE_TOTAL" "VE_FORMS" "PEDS"     "PERSONS"  "COUNTY"   "CITY"     "DAY"      "MONTH"    "YEAR"     "DAY_WEEK"
# [13] "HOUR"     "MINUTE"   "NHS"      "ROAD_FNC" "ROUTE"    "TWAY_ID"  "TWAY_ID2" "MILEPT"   "LATITUDE" "LONGITUD" "SP_JUR"   "HARM_EV" 
# [25] "MAN_COLL" "RELJCT1"  "RELJCT2"  "TYP_INT"  "REL_ROAD" "WRK_ZONE" "LGT_COND" "WEATHER1" "WEATHER2" "WEATHER"  "SCH_BUS"  "RAIL"    
# [37] "NOT_HOUR" "NOT_MIN"  "ARR_HOUR" "ARR_MIN"  "HOSP_HR"  "HOSP_MN"  "CF1"      "CF2"      "CF3"      "FATALS"   "DRUNK_DR"


#**************************
#Compare the least amount to the second least amount (2010 compared to 2011,2012,2013,2014)
#**************************
same_names <- rep(0,100)
col_names1 <- colnames(accident_data2011)
col_names2 <- colnames(accident_data2010)
for(i in 1:length(col_names1)){
  for(j in 1:length(col_names2)){
    if(col_names1[i]==col_names2[j]){
      same_names[i]=col_names2[j]
    }
  }
}
#print(subset(same_names, same_names!="0"))
same_2011_2010 <-subset(same_names, same_names!="0")
#length(same_2011_2010) #47
# [1] "STATE"    "ST_CASE"  "VE_TOTAL" "VE_FORMS" "PEDS"     "PERSONS" 
# [7] "COUNTY"   "CITY"     "DAY"      "MONTH"    "YEAR"     "DAY_WEEK"
# [13] "HOUR"     "MINUTE"   "NHS"      "ROAD_FNC" "ROUTE"    "TWAY_ID" 
# [19] "TWAY_ID2" "MILEPT"   "LATITUDE" "LONGITUD" "SP_JUR"   "HARM_EV" 
# [25] "MAN_COLL" "RELJCT1"  "RELJCT2"  "TYP_INT"  "REL_ROAD" "WRK_ZONE"
# [31] "LGT_COND" "WEATHER1" "WEATHER2" "WEATHER"  "SCH_BUS"  "RAIL"    
# [37] "NOT_HOUR" "NOT_MIN"  "ARR_HOUR" "ARR_MIN"  "HOSP_HR"  "HOSP_MN" 
# [43] "CF1"      "CF2"      "CF3"      "FATALS"   "DRUNK_DR"


#**************************
#Compare the least amount to the second least amount (2010 compared to 2015)
#**************************
same_names <- rep(0,100)
col_names1 <- colnames(accident_data2015)
col_names2 <- colnames(accident_data2010)
for(i in 1:length(col_names1)){
  for(j in 1:length(col_names2)){
    if(col_names1[i]==col_names2[j]){
      same_names[i]=col_names2[j]
    }
  }
}
#print(subset(same_names, same_names!="0"))
same_2015_2010 <-subset(same_names, same_names!="0")
#length(same_2015_2010) #46
# [1]  "STATE"    "ST_CASE"  "VE_TOTAL" "VE_FORMS" "PEDS"     "PERSONS" 
# [7]  "COUNTY"   "CITY"     "DAY"      "MONTH"    "YEAR"     "DAY_WEEK"
# [13] "HOUR"     "MINUTE"   "NHS"      "ROUTE"    "TWAY_ID"  "TWAY_ID2"
# [19] "MILEPT"   "LATITUDE" "LONGITUD" "SP_JUR"   "HARM_EV"  "MAN_COLL"
# [25] "RELJCT1"  "RELJCT2"  "TYP_INT"  "WRK_ZONE" "REL_ROAD" "LGT_COND"
# [31] "WEATHER1" "WEATHER2" "WEATHER"  "SCH_BUS"  "RAIL"     "NOT_HOUR"
# [37] "NOT_MIN"  "ARR_HOUR" "ARR_MIN"  "HOSP_HR"  "HOSP_MN"  "CF1"     
# [43] "CF2"      "CF3"      "FATALS"   "DRUNK_DR"

#**************************
#Compare the least amount to the second least amount (2011,2012,2013,2014 compared to 2015)
#**************************
same_names <- rep(0,100)
col_names1 <- colnames(accident_data2015)
col_names2 <- colnames(accident_data2011)
for(i in 1:length(col_names1)){
  for(j in 1:length(col_names2)){
    if(col_names1[i]==col_names2[j]){
      same_names[i]=col_names2[j]
    }
  }
}
#print(subset(same_names, same_names!="0"))
same_2015_2011 <-subset(same_names, same_names!="0")
#length(same_2015_2011) #49


#**************************
#(1): Compare the least amount to the second least amount (2010 compared to 2011,2012,2013,2014)
#same_2011_2010
#length(same_2011_2010) #47
#(2): Compare the least amount to the second least amount (2010 compared to 2015)
#same_2015_2010
#length(same_2015_2010) #46
#(3): Compare the least amount to the second least amount (2011,2012,2013,2014 compared to 2015)
#same_2015_2011
#length(same_2015_2011) #49
#**************************

#Compare (1) to (2)
same_names <- rep(0,100)
col_names1 <- same_2011_2010
col_names2 <- same_2015_2010
for(i in 1:length(col_names1)){
  for(j in 1:length(col_names2)){
    if(col_names1[i]==col_names2[j]){
      same_names[i]=col_names2[j]
    }
  }
}
#print(subset(same_names, same_names!="0"))
# [1]  "STATE"    "ST_CASE"  "VE_TOTAL" "VE_FORMS" "PEDS"     "PERSONS"  "COUNTY"   "CITY"     "DAY"      "MONTH"    "YEAR"    
# [12] "DAY_WEEK" "HOUR"     "MINUTE"   "NHS"      "ROUTE"    "TWAY_ID"  "TWAY_ID2" "MILEPT"   "LATITUDE" "LONGITUD" "SP_JUR"  
# [23] "HARM_EV"  "MAN_COLL" "RELJCT1"  "RELJCT2"  "TYP_INT"  "REL_ROAD" "WRK_ZONE" "LGT_COND" "WEATHER1" "WEATHER2" "WEATHER" 
# [34] "SCH_BUS"  "RAIL"     "NOT_HOUR" "NOT_MIN"  "ARR_HOUR" "ARR_MIN"  "HOSP_HR"  "HOSP_MN"  "CF1"      "CF2"      "CF3"     
# [45] "FATALS"   "DRUNK_DR"
same_1_2 <-subset(same_names, same_names!="0")
#length(same_1_2) #46


#Compare (2) to (3)
same_names <- rep(0,100)
col_names1 <- same_2011_2010
col_names2 <- same_2015_2011
for(i in 1:length(col_names1)){
  for(j in 1:length(col_names2)){
    if(col_names1[i]==col_names2[j]){
      same_names[i]=col_names2[j]
    }
  }
}
#print(subset(same_names, same_names!="0"))
#[1] "STATE"    "ST_CASE"  "VE_TOTAL" "VE_FORMS" "PEDS"     "PERSONS"  "COUNTY"   "CITY"     "DAY"      "MONTH"    "YEAR"    
#[12] "DAY_WEEK" "HOUR"     "MINUTE"   "NHS"      "ROUTE"    "TWAY_ID"  "TWAY_ID2" "MILEPT"   "LATITUDE" "LONGITUD" "SP_JUR"  
#[23] "HARM_EV"  "MAN_COLL" "RELJCT1"  "RELJCT2"  "TYP_INT"  "REL_ROAD" "WRK_ZONE" "LGT_COND" "WEATHER1" "WEATHER2" "WEATHER" 
#[34] "SCH_BUS"  "RAIL"     "NOT_HOUR" "NOT_MIN"  "ARR_HOUR" "ARR_MIN"  "HOSP_HR"  "HOSP_MN"  "CF1"      "CF2"      "CF3"     
#[45] "FATALS"   "DRUNK_DR"
same_2_3 <-subset(same_names, same_names!="0")
#length(same_2_3) #46


#Compare the comparisons
same_names <- rep(0,100)
col_names1 <- same_1_2
col_names2 <- same_2_3
for(i in 1:length(col_names1)){
  for(j in 1:length(col_names2)){
    if(col_names1[i]==col_names2[j]){
      same_names[i]=col_names2[j]
    }
  }
}
#print(subset(same_names, same_names!="0"))
# [1]  "STATE"    "ST_CASE"  "VE_TOTAL" "VE_FORMS" "PEDS"     "PERSONS"  "COUNTY"   "CITY"     "DAY"      "MONTH"    "YEAR"    
# [12] "DAY_WEEK" "HOUR"     "MINUTE"   "NHS"      "ROUTE"    "TWAY_ID"  "TWAY_ID2" "MILEPT"   "LATITUDE" "LONGITUD" "SP_JUR"  
# [23] "HARM_EV"  "MAN_COLL" "RELJCT1"  "RELJCT2"  "TYP_INT"  "REL_ROAD" "WRK_ZONE" "LGT_COND" "WEATHER1" "WEATHER2" "WEATHER" 
# [34] "SCH_BUS"  "RAIL"     "NOT_HOUR" "NOT_MIN"  "ARR_HOUR" "ARR_MIN"  "HOSP_HR"  "HOSP_MN"  "CF1"      "CF2"      "CF3"     
# [45] "FATALS"   "DRUNK_DR"


#*******************************************************************************
#1st: Compare the ones that are not the SAME Variables that are in each dataset
#*******************************************************************************
#setdiff(colnames(accident_data2015), colnames(accident_data2011))
#[1] "RUR_URB"  "FUNC_SYS" "RD_OWNER"
#setdiff(colnames(accident_data2011), colnames(accident_data2015))
#[1] "ROAD_FNC"


#setdiff(colnames(accident_data2015), colnames(accident_data2010))
#[1] "PVH_INVL"   "PERNOTMVIT" "PERMVIT"    "RUR_URB"    "FUNC_SYS"   "RD_OWNER" 
#setdiff(colnames(accident_data2010), colnames(accident_data2015))
#[1] "ROAD_FNC"

#setdiff(colnames(accident_data2010), colnames(accident_data2011))
#0
#setdiff(colnames(accident_data2011), colnames(accident_data2010))
#[1] "PVH_INVL"   "PERNOTMVIT" "PERMVIT" 

#*******************************************************************************
#RANDOM SUMMARIES
#sort(unique(accident_data2012$PVH_INVL)) # 0 1 2 3 4 5 6 9
#sort(unique(accident_data2013$PVH_INVL)) # 0 1 2 3 4 5 6 7 8 11
#sort(unique(accident_data2014$PVH_INVL)) # 0 1 2 3 4 5 6 7 10 11
#*******************************************************************************


#**********************************************************************************
#Decision: Need to remove all colnames from:
# "PVH_INVL", "PERNOTMVIT", "PERMVIT", "RUR_URB", "FUNC_SYS", "RD_OWNER", "ROAD_FNC"
#This will produce datatsets with consistent variables in them.
#**********************************************************************************

# ***** 2015 *****
vect_position <- rep(0,length(accident_data2015))
for(i in 1:length(accident_data2015)){
  if(colnames(accident_data2015)[i] == "PVH_INVL"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2015)[i] == "PERNOTMVIT"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2015)[i] == "PERMVIT"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2015)[i] == "RUR_URB"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2015)[i] == "FUNC_SYS"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2015)[i] == "RD_OWNER"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2015)[i] == "ROAD_FNC"){
    vect_position[i] <- i
  }
}
#vect_position
vect_position <- subset(vect_position, vect_position!=0)
#vect_position
subset_accident_data2015 <- accident_data2015[,-c(vect_position)]
#dim(subset_accident_data2015) #32166  46
#View(subset_accident_data2015)


# ***** 2014 *****
vect_position <- rep(0,length(accident_data2014))
for(i in 1:length(accident_data2014)){
  if(colnames(accident_data2014)[i] == "PVH_INVL"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2014)[i] == "PERNOTMVIT"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2014)[i] == "PERMVIT"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2014)[i] == "RUR_URB"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2014)[i] == "FUNC_SYS"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2014)[i] == "RD_OWNER"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2014)[i] == "ROAD_FNC"){
    vect_position[i] <- i
  }
}
#vect_position
vect_position <- subset(vect_position, vect_position!=0)
#vect_position
subset_accident_data2014 <- accident_data2014[,-c(vect_position)]
#dim(subset_accident_data2014) #30056    46
#View(subset_accident_data2014)


# ***** 2013 *****
vect_position <- rep(0,length(accident_data2013))
for(i in 1:length(accident_data2013)){
  if(colnames(accident_data2013)[i] == "PVH_INVL"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2013)[i] == "PERNOTMVIT"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2013)[i] == "PERMVIT"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2013)[i] == "RUR_URB"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2013)[i] == "FUNC_SYS"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2013)[i] == "RD_OWNER"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2013)[i] == "ROAD_FNC"){
    vect_position[i] <- i
  }
}
#vect_position
vect_position <- subset(vect_position, vect_position!=0)
#vect_position
subset_accident_data2013 <- accident_data2013[,-c(vect_position)]
#dim(subset_accident_data2013)  #30202    46
#View(subset_accident_data2013)


# ***** 2012 *****
vect_position <- rep(0,length(accident_data2012))
for(i in 1:length(accident_data2012)){
  if(colnames(accident_data2012)[i] == "PVH_INVL"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2012)[i] == "PERNOTMVIT"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2012)[i] == "PERMVIT"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2012)[i] == "RUR_URB"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2012)[i] == "FUNC_SYS"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2012)[i] == "RD_OWNER"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2012)[i] == "ROAD_FNC"){
    vect_position[i] <- i
  }
}
#vect_position
vect_position <- subset(vect_position, vect_position!=0)
#vect_position
subset_accident_data2012 <- accident_data2012[,-c(vect_position)]
#dim(subset_accident_data2012)  #31006    46
#View(subset_accident_data2012)


# ***** 2011 *****
vect_position <- rep(0,length(accident_data2011))
for(i in 1:length(accident_data2011)){
  if(colnames(accident_data2011)[i] == "PVH_INVL"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2011)[i] == "PERNOTMVIT"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2011)[i] == "PERMVIT"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2011)[i] == "RUR_URB"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2011)[i] == "FUNC_SYS"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2011)[i] == "RD_OWNER"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2011)[i] == "ROAD_FNC"){
    vect_position[i] <- i
  }
}
#vect_position
vect_position <- subset(vect_position, vect_position!=0)
#vect_position
subset_accident_data2011 <- accident_data2011[,-c(vect_position)]
#dim(subset_accident_data2011) #29867    46
#View(subset_accident_data2011)


# ***** 2010 *****
vect_position <- rep(0,length(accident_data2010))
for(i in 1:length(accident_data2010)){
  if(colnames(accident_data2010)[i] == "PVH_INVL"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2010)[i] == "PERNOTMVIT"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2010)[i] == "PERMVIT"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2010)[i] == "RUR_URB"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2010)[i] == "FUNC_SYS"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2010)[i] == "RD_OWNER"){
    vect_position[i] <- i
  }
  if(colnames(accident_data2010)[i] == "ROAD_FNC"){
    vect_position[i] <- i
  }
}
#vect_position
vect_position <- subset(vect_position, vect_position!=0)
#vect_position
subset_accident_data2010 <- accident_data2010[,-c(vect_position)]
dim(subset_accident_data2010) #30296    46
#View(subset_accident_data2011)


#if(sum(vect_position)==0) {subset_accident = accident_data2010}
#if(sum(vect_position)!=0) {subset_accident_data2010 = subset(vect_position, vect_position!=0)}
#vect_position <- subset(vect_position, vect_position!=0)
#vect_position
#if(sum(vect_position)!=0) {subset_accident_data2010 <- accident_data2010[,-c(vect_position)]}
#dim(subset_accident)  #30296    47
#dim(subset_accident_data2010)
#dim(accident_data2010) 


#**********************************************************************************************************************
# MERGING all datasets with like columns ------- EXAMPLE
#**********************************************************************************************************************
#Small example to see how this is done!
#Notice: matrix did not work before because a matrix has to be a diagonal matrix (same amount of columns as names)
#Instead: use dataframes
#In order to use rowbind 'rbind' the variables names that are being combined need to be the EXACT same
a <- c(1,2,3,4,5,6)
b <- c(7,8,9,10,11,12)
mat <- data.frame(a,b)
#mat
#dim(mat)
#View(mat)
d <-c(13,14,15,16,17,18)
e <-c(1,1,1,1,1,1)
mat_2 <- data.frame(d,e)
colnames(mat_2) <-c("a","b")
#dim(mat_2)
new_mat <- rbind(mat,mat_2)
#dim(new_mat)
#View(new_mat)

#setdiff(colnames(subset_accident_data2015), colnames(subset_accident_data2014))
#setdiff(colnames(subset_accident_data2014), colnames(subset_accident_data2015))
#[1] "ROAD_FNC"

#***************************************************************
# ROWBIND all datasets with like columns
#***************************************************************

all_merged_accidents <- rbind(subset_accident_data2015, subset_accident_data2014, subset_accident_data2013, subset_accident_data2012, subset_accident_data2011, subset_accident_data2010)
#View(all_merged_accidents)


#*****************************************************************************************************************************
#Look at code to plot the exploratory analysis plot observing Accidents, Fatalities, & DUIs across 2015.
#*****************************************************************************************************************************
#Create the merged data ste
#Create a unique ID number for each case; This is important because each year repeats the case numbers used in each state
uniqueID <- seq(1:183593)
all_merged_accidents$uniqueID <- uniqueID
#View(all_merged_accidents)

#Create a state accident ID that has a year extension
newST_CASE <- rep(0,183593)
for(i in 1:183593){
  if(all_merged_accidents$YEAR[i]==2010){
    newST_CASE[i] <- paste(all_merged_accidents$ST_CASE[i],-2010,sep="")
  }
  if(all_merged_accidents$YEAR[i]==2011){
    newST_CASE[i] <- paste(all_merged_accidents$ST_CASE[i],-2011,sep="")
  }
  if(all_merged_accidents$YEAR[i]==2012){
    newST_CASE[i] <- paste(all_merged_accidents$ST_CASE[i],-2012,sep="")
  }
  if(all_merged_accidents$YEAR[i]==2013){
    newST_CASE[i] <- paste(all_merged_accidents$ST_CASE[i],-2013,sep="")
  }
  if(all_merged_accidents$YEAR[i]==2014){
    newST_CASE[i] <- paste(all_merged_accidents$ST_CASE[i],-2014,sep="")
  }
  if(all_merged_accidents$YEAR[i]==2015){
    newST_CASE[i] <- paste(all_merged_accidents$ST_CASE[i],-2015,sep="")
  }
}
#length(newST_CASE) #183593
#nrow(all_merged_accidents) #183593
#Add this variable to the merged dataset
all_merged_accidents$newST_CASE <- newST_CASE

#************************************************************************
tot_days_endofeachmonth <- c(31,59,90,120,151,181,212,243,273,304,334,365)
max<-c(0,31,28,31,30,31,30,31,31,30,31,30) 
#length(max) #12

#RECODE day so that days go from 1-365 starting from January 1st 2015 to December 31st 2015
day<-rep(0,nrow(accident_data2015))
for(i in 1:nrow(accident_data2015)){
  if(accident_data2015$YEAR[i]==2015){
    if(accident_data2015$MONTH[i]==1){
      day[i]<-max[1] + accident_data2015$DAY[i]
    }
    if(accident_data2015$MONTH[i]==2){
      day[i]<-max[1] + max[2] + accident_data2015$DAY[i]
    }
    if(accident_data2015$MONTH[i]==3){
      day[i]<-max[1] + max[2] + max[3] + accident_data2015$DAY[i]
    }
    if(accident_data2015$MONTH[i]==4){
      day[i]<-max[1] + max[2] + max[3] + max[4] + accident_data2015$DAY[i]
    }
    if(accident_data2015$MONTH[i]==5){
      day[i]<-max[1] + max[2] + max[3] + max[4] + max[5] + accident_data2015$DAY[i]
    }
    if(accident_data2015$MONTH[i]==6){
      day[i]<-max[1] + max[2] + max[3] + max[4] + max[5] + max[6] + accident_data2015$DAY[i]
    }
    if(accident_data2015$MONTH[i]==7){
      day[i]<-max[1] + max[2] + max[3] + max[4] + max[5] + max[6] + max[7] + accident_data2015$DAY[i]
    }
    if(accident_data2015$MONTH[i]==8){
      day[i]<-max[1] + max[2] + max[3] + max[4] + max[5] + max[6] + max[7] + max[8] + accident_data2015$DAY[i]
    }
    if(accident_data2015$MONTH[i]==9){
      day[i]<-max[1] + max[2] + max[3] + max[4] + max[5] + max[6] + max[7] + max[8] + max[9] + accident_data2015$DAY[i]
    }
    if(accident_data2015$MONTH[i]==10){
      day[i]<-max[1] + max[2] + max[3] + max[4] + max[5] + max[6] + max[7] + max[8] + max[9] + max[10] + accident_data2015$DAY[i]
    }
    if(accident_data2015$MONTH[i]==11){
      day[i]<-max[1] + max[2] + max[3] + max[4] + max[5] + max[6] + max[7] + max[8] + max[9] + max[10] + max[11] + accident_data2015$DAY[i]
    }
    if(accident_data2015$MONTH[i]==12){
      day[i]<-max[1] + max[2] + max[3] + max[4] + max[5] + max[6] + max[7] + max[8] + max[9] + max[10] + max[11] + max[12] + accident_data2015$DAY[i]
    }
  }
}  
#day
#sort(unique(day))


#**************
#Calculate the number of fatalities on each day in each state in 2015
#**************
fatal<-rep(0,56)
for(i in 1:nrow(accident_data2015)){
  for(j in 1:56){
    if(accident_data2015$STATE[i]==j)
      fatal[j] <- fatal[j] + accident_data2015$FATALS[i]
    
  }
}  
#fatal
Tot_Fatal_15 <- fatal
state<-1:56
df1<-data.frame(fatal,state)
df1[-c(3,7,14,43,52),]
#sum(accident_data2015$FATALS) #Check to make sure the sum is accurate

#**************
#Calculate the number of Accidents on each day in each state in 2015
#**************
total<-rep(0,56)
for(i in 1:nrow(accident_data2015)){
  for(j in 1:56){
    if(accident_data2015$STATE[i]==j){
      total[j]<-total[j]+1
    }
  }
}
#total
#Total_Accident_15 <- accident

state<-1:56
df5<-data.frame(total,state)
df5[-c(3,7,14,43,52),]

#**************
#Calculate the number of DUIs on each day for each state in 2015
#**************
accident_data2015$DRUNK_DR<-ifelse(accident_data2015$DRUNK_DR>=1,1,0)
dui<-rep(0,56)
for(i in 1:nrow(accident_data2015)){
  for(j in 1:56){
    if(accident_data2015$STATE[i]==j)
      dui[j] <- dui[j] + accident_data2015$DRUNK_DR[i]
    
  }
}
#dui
#sum(dui) #8862
#length(which(accident_data2015$DRUNK_DR==1)) #8862

state<-1:56
df2<-data.frame(dui,state)
df2[-c(3,7,14,43,52),]
df2[-c(3,7,14,43,52),]
#Tot_DUI_15 <- dui


#*****************************************************************************************************************************
#Look at code to plot the exploratory analysis plot observing Accidents, Fatalities, & DUIs across 2010-2015.
#**********************************************************************************************************************************
#Evaluate the number of accidents and the number of fatalities that have occured between 2010-2015
max<-c(0,31,28,31,30,31,30,31,31,30,31,30) 
max_LP<-c(0,31,29,31,30,31,30,31,31,30,31,30) 
max_year<-c(0,365,730,1096,1461,1826)
#max_year
#This tries to look at a for loop and account for the leap year afterwards
#all_merged_accidents
day<-rep(0,nrow(all_merged_accidents))
for(i in 1:nrow(all_merged_accidents)){
  for(j in 1:6){
    if(all_merged_accidents$YEAR[i]==j+2009){
      max <-c(0,31,28,31,30,31,30,31,31,30,31,30) 
      if(all_merged_accidents$YEAR[i]==2012){
        max=max_LP
      }
      if(all_merged_accidents$MONTH[i]==1){
        day[i]<-max_year[j] + max[1] + all_merged_accidents$DAY[i]
      }
      if(all_merged_accidents$MONTH[i]==2){
        day[i]<-max_year[j] + max[1] + max[2] + all_merged_accidents$DAY[i]
      }
      if(all_merged_accidents$MONTH[i]==3){
        day[i]<-max_year[j] + max[1] + max[2] + max[3] + all_merged_accidents$DAY[i]
      }
      if(all_merged_accidents$MONTH[i]==4){
        day[i]<-max_year[j] + max[1] + max[2] + max[3] + max[4] + all_merged_accidents$DAY[i]
      }
      if(all_merged_accidents$MONTH[i]==5){
        day[i]<-max_year[j] + max[1] + max[2] + max[3] + max[4] + max[5] + all_merged_accidents$DAY[i]
      }
      if(all_merged_accidents$MONTH[i]==6){
        day[i]<-max_year[j] + max[1] + max[2] + max[3] + max[4] + max[5] + max[6] + all_merged_accidents$DAY[i]
      }
      if(all_merged_accidents$MONTH[i]==7){
        day[i]<-max_year[j] + max[1] + max[2] + max[3] + max[4] + max[5] + max[6] + max[7] + all_merged_accidents$DAY[i]
      }
      if(all_merged_accidents$MONTH[i]==8){
        day[i]<-max_year[j] + max[1] + max[2] + max[3] + max[4] + max[5] + max[6] + max[7] + max[8] + all_merged_accidents$DAY[i]
      }
      if(all_merged_accidents$MONTH[i]==9){
        day[i]<-max_year[j] + max[1] + max[2] + max[3] + max[4] + max[5] + max[6] + max[7] + max[8] + max[9] + all_merged_accidents$DAY[i]
      }
      if(all_merged_accidents$MONTH[i]==10){
        day[i]<-max_year[j] + max[1] + max[2] + max[3] + max[4] + max[5] + max[6] + max[7] + max[8] + max[9] + max[10] + all_merged_accidents$DAY[i]
      }
      if(all_merged_accidents$MONTH[i]==11){
        day[i]<-max_year[j] + max[1] + max[2] + max[3] + max[4] + max[5] + max[6] + max[7] + max[8] + max[9] + max[10] + max[11] + all_merged_accidents$DAY[i]
      }
      if(all_merged_accidents$MONTH[i]==12){
        day[i]<-max_year[j] + max[1] + max[2] + max[3] + max[4] + max[5] + max[6] + max[7] + max[8] + max[9] + max[10] + max[11] + max[12] + all_merged_accidents$DAY[i]
      }
    }
  }
}  
#sort(unique(day))
all_merged_accidents$day <- day

#**************
#Calculate the number of accidents occuring on the same day through 5 years being observed
#**************
#Find out how many accidents are occuring on the same day
#length(unique(day)) #2191
total<-rep(0,2191)
for(j in 1:2191){
  total[j]<-length(which(all_merged_accidents$day==j))
}
#total
#sum(total) #Check to make sure this is correct
#nrow(all_merged_accidents) #Verify that the number of accident files matches 
day_number<-1:2191
df_numb_yearly_accidents <- data.frame(day_number, total) 
#View(df_numb_yearly_accidents)

#Check
#Check Day 366-- represents the 1st day of year 2011
#length(which(all_merged_accidents$YEAR==2011 & all_merged_accidents$MONTH==1 & all_merged_accidents$DAY==1)) #117
#Check Day 1461-- represents the last day of the year 2013
#length(which(all_merged_accidents$YEAR==2013 & all_merged_accidents$MONTH==12 & all_merged_accidents$DAY==31)) #77

#Plot Total vs Day
#length(day)
#plot(day_number, total,
#     xlab="Day Number (01/01/10-12/31/15)",
#     ylab="Total Number of Accidents",
#     main="Accident Fatality Seasonality",
#     col="blue")
#abline(v=166,col="red")
#abline(v=258,col="red")
#abline(v=151,col="red")
#abline(v=273,col="red")
#abline(v=151+365,col="red")
#abline(v=273+365,col="red")
#abline(v=151+365+366,col="red")
#abline(v=273+365+366,col="red")
#abline(v=151+365+366+365,col="red")
#bline(v=273+365+366+365,col="red")
#abline(v=151+365+366+365+365,col="red")
#abline(v=273+365+366+365+365,col="red")
#abline(v=151+365+366+365+365+365,col="red")
#abline(v=273+365+366+365+365+365,col="red")

#************
#Calculate the number of fatalities that occur on the same day from 2010-2015
#************
count<-rep(0,2191)
for(i in 1:nrow(all_merged_accidents)){
  for(j in 1:2191){
    if(day[i]==j){
      count[j]<-count[j]+all_merged_accidents$FATALS[i]
    }
  }
}
#count
#These values should be equal
#sum(count) #199989
#sum(all_merged_accidents$FATALS) #199989

#Look at plots
#plot(day_number, count)
#boxplot(day_number, count)
#hist(count)

#************
#Counting the number of DUIs that occur on every day
#************
#First make the DUI variable dicotomous
#summary(all_merged_accidents$DRUNK_DR) 
##   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
#0.00    0.00    0.00    0.31    1.00    4.00
all_merged_accidents$DRUNK_DR<-ifelse(all_merged_accidents$DRUNK_DR>=1,1,0)
#summary(all_merged_accidents$DRUNK_DR)
#Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
#0.000   0.000   0.000   0.301   1.000   1.000

#Count the Number of Drunk Drivers
#length(which(all_merged_accidents$DRUNK_DR==1)) #5592
drunk<-rep(0,2191)
for(i in 1:nrow(all_merged_accidents)){
  for(j in 1:2191){
    if(day[i]==j){
      drunk[j]<-drunk[j]+all_merged_accidents$DRUNK_DR[i]
    }
  }
}
#drunk
#These values should be equal
#sum(drunk) #55262
#sum(all_merged_accidents$DRUNK_DR) #55262


#************************************************
#PLOTTING ALL THREE PLOTS
#Create plots on top of one another
par(mfrow=c(3,1))
#************************************************
#First-- number of accidents on a given day
plot(day_number, total,
     xlab="Day Number (01/01/10-12/31/15)",
     ylab="Total Number of Accidents",
     main="Total Number of Motor Vehicle Related Accidents (Resulting in a Fatality) on Each Day Between 2010-2015",
     col="blue")
lines(lowess(total ~ day_number, f = 0.08), col = "black", lwd = 2)
#Account for the summer months
abline(v=151,col="red")
abline(v=273,col="red")
abline(v=151+365,col="red")
abline(v=273+365,col="red")
abline(v=151+365+366,col="red")
abline(v=273+365+366,col="red")
abline(v=151+365+366+365,col="red")
abline(v=273+365+366+365,col="red")
abline(v=151+365+366+365+365,col="red")
abline(v=273+365+366+365+365,col="red")
abline(v=151+365+366+365+365+365,col="red")
abline(v=273+365+366+365+365+365,col="red")

#Evaluate NYEs 12/31-1/1 each year
#2010 1/1 -- day1
#2010 12/31-- day365
#NYE_day<-c(365,365+365, 366+(365*2), 366+(365*3), 366+(365*4), 366+(365*5))
#NY_day<-c(1,1+365, 1+(365*2), 1+366+(365*2), 1+366+(365*3), 1+366+(365*4))
#NYE_day
#NY_day
abline(v=1,col="cyan")
abline(v=365,col="cyan")
abline(v=366,col="cyan")
abline(v=730,col="cyan")
abline(v=731,col="cyan")
abline(v=1096,col="cyan")
abline(v=1097,col="cyan")
abline(v=1461,col="cyan")
abline(v=1462,col="cyan")
abline(v=1826,col="cyan")
abline(v=1827,col="cyan")
abline(v=1461,col="cyan")
abline(v=2191,col="cyan")

#Evaluate 4th of July
#2010 7/4 -- day185
#2010 7/5-- day186
#July4_before<-c(183,183+365, 183+(365+366), 183+366+(365*2), 183+366+(365*3), 183+366+(365*4))
#July4_day<-c(185,185+365, 185+(365+366), 185+366+(365*2), 185+366+(365*3), 185+366+(365*4))
#July4_dayafter<-c(186,186+365, 186+(365+366), 186+366+(365*2), 186+366+(365*3), 186+366+(365*4))
#July4_before
#July4_day
#July4_dayafter
abline(v=183,col="purple")
abline(v=186,col="purple")
abline(v=548,col="purple")
abline(v=551,col="purple")
abline(v=914,col="purple")
abline(v=917,col="purple")
abline(v=1279,col="purple")
abline(v=1282,col="purple")
abline(v=1644,col="purple")
abline(v=1647,col="purple")
abline(v=2009,col="purple")
abline(v=2012,col="purple")


#Evaluate Thanksgiving
#2010 11/24 -- day328
#2010 11/25 -- day329
#2010 11/26-- day330
#TGeve<-c(328,328+365, 328+(365+366), 328+366+(365*2), 328+366+(365*3), 328+366+(365*4))
#TGdayafter<-c(330,330+365, 330+(365+366), 330+366+(365*2), 330+366+(365*3), 330+366+(365*4))
#TGeve
#TGdayafter
abline(v=328,col="yellow")
abline(v=330,col="yellow")
abline(v=693,col="yellow")
abline(v=695,col="yellow")
abline(v=1059,col="yellow")
abline(v=1061,col="yellow")
abline(v=1424,col="yellow")
abline(v=1426,col="yellow")
abline(v=1789,col="yellow")
abline(v=1791,col="yellow")
abline(v=2154,col="yellow")
abline(v=2156,col="yellow")


#Evaluate Christmas
#2010 12/24 -- day358
#2010 11/25 -- day359
#2010 11/26-- day360
#XmanEve<-c(358,358+365, 358+(365+366), 358+366+(365*2), 358+366+(365*3), 358+366+(365*4))
#XmasAfter<-c(360,360+365, 360+(365+366), 360+366+(365*2), 360+366+(365*3), 360+366+(365*4))
#XmanEve
#XmasAfter
abline(v=358,col="green")
abline(v=360,col="green")
abline(v=723,col="green")
abline(v=725,col="green")
abline(v=1089,col="green")
abline(v=1091,col="green")
abline(v=1454,col="green")
abline(v=1456,col="green")
abline(v=1819,col="green")
abline(v=1821,col="green")
abline(v=2184,col="green")
abline(v=2186,col="green")

#Evaluate Labor Day
#2010 9/6 -- day249
#2010 9/7 -- day250
#LDBefore<-c(247,247+365, 247+(365+366), 247+366+(365*2), 247+366+(365*3), 247+366+(365*4))
#LD<-c(249,249+365, 249+(365+366), 249+366+(365*2), 249+366+(365*3), 249+366+(365*4))
#LDAfter<-c(250,250+365, 250+(365+366), 250+366+(365*2), 250+366+(365*3), 250+366+(365*4))
#LDBefore
#LD
#LDAfter
abline(v=247,col="pink")
abline(v=250,col="pink")
abline(v=612,col="pink")
abline(v=615,col="pink")
abline(v=978,col="pink")
abline(v=981,col="pink")
abline(v=1343,col="pink")
abline(v=1346,col="pink")
abline(v=1708,col="pink")
abline(v=1711,col="pink")
abline(v=2073,col="pink")
abline(v=2076,col="pink")
data<-data.frame(NYE_day,July4_before,July4_day,LDAfter,TGeve, XmanEve)
names<-c("NYE", "SummerInterval", "July4th", "LaborDay", "Thnksgvng", "Xmas")
legend('topleft', names(data), 
       lty=1, col=c('cyan', 'red', 'purple','pink','yellow', 'green'), bty='n', cex=.75)


#************************************************
#Second-- number of fatalities on a given day
plot(day_number, count,
     xlab="Day Number (01/01/10-12/31/15)",
     ylab="Total Number of Fatalities",
     main="Total Motor Vehicle Related Fatalities on Each Day Between 2010-2015",
     col="blue")
lines(lowess(count ~ day_number, f = 0.08), col = "black", lwd = 2)
#Account for the summer months
abline(v=151,col="red")
abline(v=273,col="red")
abline(v=151+365,col="red")
abline(v=273+365,col="red")
abline(v=151+365+366,col="red")
abline(v=273+365+366,col="red")
abline(v=151+365+366+365,col="red")
abline(v=273+365+366+365,col="red")
abline(v=151+365+366+365+365,col="red")
abline(v=273+365+366+365+365,col="red")
abline(v=151+365+366+365+365+365,col="red")
abline(v=273+365+366+365+365+365,col="red")

#Evaluate NYEs 12/31-1/1 each year
#2010 1/1 -- day1
#2010 12/31-- day365
#NYE_day<-c(365,365+365, 366+(365*2), 366+(365*3), 366+(365*4), 366+(365*5))
#NY_day<-c(1,1+365, 1+(365*2), 1+366+(365*2), 1+366+(365*3), 1+366+(365*4))
#NYE_day
#NY_day
abline(v=1,col="cyan")
abline(v=365,col="cyan")
abline(v=366,col="cyan")
abline(v=730,col="cyan")
abline(v=731,col="cyan")
abline(v=1096,col="cyan")
abline(v=1097,col="cyan")
abline(v=1461,col="cyan")
abline(v=1462,col="cyan")
abline(v=1826,col="cyan")
abline(v=1827,col="cyan")
abline(v=1461,col="cyan")
abline(v=2191,col="cyan")

#Evaluate 4th of July
#2010 7/4 -- day185
#2010 7/5-- day186
#July4_before<-c(183,183+365, 183+(365+366), 183+366+(365*2), 183+366+(365*3), 183+366+(365*4))
#July4_day<-c(185,185+365, 185+(365+366), 185+366+(365*2), 185+366+(365*3), 185+366+(365*4))
#July4_dayafter<-c(186,186+365, 186+(365+366), 186+366+(365*2), 186+366+(365*3), 186+366+(365*4))
#July4_before
#July4_day
#July4_dayafter
abline(v=183,col="purple")
abline(v=186,col="purple")
abline(v=548,col="purple")
abline(v=551,col="purple")
abline(v=914,col="purple")
abline(v=917,col="purple")
abline(v=1279,col="purple")
abline(v=1282,col="purple")
abline(v=1644,col="purple")
abline(v=1647,col="purple")
abline(v=2009,col="purple")
abline(v=2012,col="purple")


#Evaluate Thanksgiving
#2010 11/24 -- day328
#2010 11/25 -- day329
#2010 11/26-- day330
#TGeve<-c(328,328+365, 328+(365+366), 328+366+(365*2), 328+366+(365*3), 328+366+(365*4))
#TGdayafter<-c(330,330+365, 330+(365+366), 330+366+(365*2), 330+366+(365*3), 330+366+(365*4))
#TGeve
#TGdayafter
abline(v=328,col="yellow")
abline(v=330,col="yellow")
abline(v=693,col="yellow")
abline(v=695,col="yellow")
abline(v=1059,col="yellow")
abline(v=1061,col="yellow")
abline(v=1424,col="yellow")
abline(v=1426,col="yellow")
abline(v=1789,col="yellow")
abline(v=1791,col="yellow")
abline(v=2154,col="yellow")
abline(v=2156,col="yellow")


#Evaluate Christmas
#2010 12/24 -- day358
#2010 11/25 -- day359
#2010 11/26-- day360
#XmanEve<-c(358,358+365, 358+(365+366), 358+366+(365*2), 358+366+(365*3), 358+366+(365*4))
#XmasAfter<-c(360,360+365, 360+(365+366), 360+366+(365*2), 360+366+(365*3), 360+366+(365*4))
#XmanEve
#XmasAfter
abline(v=358,col="green")
abline(v=360,col="green")
abline(v=723,col="green")
abline(v=725,col="green")
abline(v=1089,col="green")
abline(v=1091,col="green")
abline(v=1454,col="green")
abline(v=1456,col="green")
abline(v=1819,col="green")
abline(v=1821,col="green")
abline(v=2184,col="green")
abline(v=2186,col="green")

#Evaluate Labor Day
#2010 9/6 -- day249
#2010 9/7 -- day250
#LDBefore<-c(247,247+365, 247+(365+366), 247+366+(365*2), 247+366+(365*3), 247+366+(365*4))
#LD<-c(249,249+365, 249+(365+366), 249+366+(365*2), 249+366+(365*3), 249+366+(365*4))
#LDAfter<-c(250,250+365, 250+(365+366), 250+366+(365*2), 250+366+(365*3), 250+366+(365*4))
#LDBefore
#LD
#LDAfter
abline(v=247,col="pink")
abline(v=250,col="pink")
abline(v=612,col="pink")
abline(v=615,col="pink")
abline(v=978,col="pink")
abline(v=981,col="pink")
abline(v=1343,col="pink")
abline(v=1346,col="pink")
abline(v=1708,col="pink")
abline(v=1711,col="pink")
abline(v=2073,col="pink")
abline(v=2076,col="pink")
data<-data.frame(NYE_day,July4_before,July4_day,LDAfter,TGeve, XmanEve)
names<-c("NYE", "SummerInterval", "July4th", "LaborDay", "Thnksgvng", "Xmas")
legend('topleft', names(data), 
       lty=1, col=c('cyan', 'red', 'purple','pink','yellow', 'green'), bty='n', cex=.75)


#************************************************
#THIRD
plot(day_number, drunk,
     xlab="Day Number (01/01/10-12/31/15)",
     ylab="Total Number of Drunk Drivers",
     main="Total Motor Vehicle Related Fatalities Involved by Drunk Drivers on Each Day Between 2010-2015",
     col="blue")
lines(lowess(drunk ~ day_number, f = 0.08), col = "black", lwd = 2)
#Account for the summer months
abline(v=151,col="red")
abline(v=273,col="red")
abline(v=151+365,col="red")
abline(v=273+365,col="red")
abline(v=151+365+366,col="red")
abline(v=273+365+366,col="red")
abline(v=151+365+366+365,col="red")
abline(v=273+365+366+365,col="red")
abline(v=151+365+366+365+365,col="red")
abline(v=273+365+366+365+365,col="red")
abline(v=151+365+366+365+365+365,col="red")
abline(v=273+365+366+365+365+365,col="red")

#Evaluate NYEs 12/31-1/1 each year
#2010 1/1 -- day1
#2010 12/31-- day365
#NYE_day<-c(365,365+365, 366+(365*2), 366+(365*3), 366+(365*4), 366+(365*5))
#NY_day<-c(1,1+365, 1+(365*2), 1+366+(365*2), 1+366+(365*3), 1+366+(365*4))
#NYE_day
#NY_day
abline(v=1,col="cyan")
abline(v=365,col="cyan")
abline(v=366,col="cyan")
abline(v=730,col="cyan")
abline(v=731,col="cyan")
abline(v=1096,col="cyan")
abline(v=1097,col="cyan")
abline(v=1461,col="cyan")
abline(v=1462,col="cyan")
abline(v=1826,col="cyan")
abline(v=1827,col="cyan")
abline(v=1461,col="cyan")
abline(v=2191,col="cyan")

#Evaluate 4th of July
#2010 7/4 -- day185
#2010 7/5-- day186
#July4_before<-c(183,183+365, 183+(365+366), 183+366+(365*2), 183+366+(365*3), 183+366+(365*4))
#July4_day<-c(185,185+365, 185+(365+366), 185+366+(365*2), 185+366+(365*3), 185+366+(365*4))
#July4_dayafter<-c(186,186+365, 186+(365+366), 186+366+(365*2), 186+366+(365*3), 186+366+(365*4))
#July4_before
#July4_day
#July4_dayafter
abline(v=183,col="purple")
abline(v=186,col="purple")
abline(v=548,col="purple")
abline(v=551,col="purple")
abline(v=914,col="purple")
abline(v=917,col="purple")
abline(v=1279,col="purple")
abline(v=1282,col="purple")
abline(v=1644,col="purple")
abline(v=1647,col="purple")
abline(v=2009,col="purple")
abline(v=2012,col="purple")


#Evaluate Thanksgiving
#2010 11/24 -- day328
#2010 11/25 -- day329
#2010 11/26-- day330
#TGeve<-c(328,328+365, 328+(365+366), 328+366+(365*2), 328+366+(365*3), 328+366+(365*4))
#TGdayafter<-c(330,330+365, 330+(365+366), 330+366+(365*2), 330+366+(365*3), 330+366+(365*4))
#TGeve
#TGdayafter
abline(v=328,col="yellow")
abline(v=330,col="yellow")
abline(v=693,col="yellow")
abline(v=695,col="yellow")
abline(v=1059,col="yellow")
abline(v=1061,col="yellow")
abline(v=1424,col="yellow")
abline(v=1426,col="yellow")
abline(v=1789,col="yellow")
abline(v=1791,col="yellow")
abline(v=2154,col="yellow")
abline(v=2156,col="yellow")


#Evaluate Christmas
#2010 12/24 -- day358
#2010 11/25 -- day359
#2010 11/26-- day360
#XmanEve<-c(358,358+365, 358+(365+366), 358+366+(365*2), 358+366+(365*3), 358+366+(365*4))
#XmasAfter<-c(360,360+365, 360+(365+366), 360+366+(365*2), 360+366+(365*3), 360+366+(365*4))
#XmanEve
#XmasAfter
abline(v=358,col="green")
abline(v=360,col="green")
abline(v=723,col="green")
abline(v=725,col="green")
abline(v=1089,col="green")
abline(v=1091,col="green")
abline(v=1454,col="green")
abline(v=1456,col="green")
abline(v=1819,col="green")
abline(v=1821,col="green")
abline(v=2184,col="green")
abline(v=2186,col="green")

#Evaluate Labor Day
#2010 9/6 -- day249
#2010 9/7 -- day250
#LDBefore<-c(247,247+365, 247+(365+366), 247+366+(365*2), 247+366+(365*3), 247+366+(365*4))
#LD<-c(249,249+365, 249+(365+366), 249+366+(365*2), 249+366+(365*3), 249+366+(365*4))
#LDAfter<-c(250,250+365, 250+(365+366), 250+366+(365*2), 250+366+(365*3), 250+366+(365*4))
#LDBefore
#LD
#LDAfter
abline(v=247,col="pink")
abline(v=250,col="pink")
abline(v=612,col="pink")
abline(v=615,col="pink")
abline(v=978,col="pink")
abline(v=981,col="pink")
abline(v=1343,col="pink")
abline(v=1346,col="pink")
abline(v=1708,col="pink")
abline(v=1711,col="pink")
abline(v=2073,col="pink")
abline(v=2076,col="pink")

data<-data.frame(NYE_day,July4_before,July4_day,LDAfter,TGeve, XmanEve)
names<-c("NYE", "SummerInterval", "July4th", "LaborDay", "Thnksgvng", "Xmas")
colnames(data)<-names
legend('topleft', names(data), 
       lty=1, col=c('cyan', 'red', 'purple','pink','yellow', 'green'), bty='n', cex=.75)
```

##III. Methods
The method that I used in this analysis is a Poisson regression model, specifically a QuasiPoisson regression model.  My outcome is defined as the total number of fatalities for each state in 2015.  Due to the fact that Poisson regression requires a response variable to be composed of count data, this condition of the outcome variable is accounted for.  Furthermore, I was suspicious the outcome variable did not adhere to the Poisson assumptions, in which the mean and variance of the outcome variable are equal.  Thus, I used the Quasipoisson regression, which accounts for over-dispersion/under-dispersion if it exists.  

First, although the outcome of interest is annual state fatality counts per state, since each state varies in population the fatality counts need to be standardized.  Although there are many ways to potentially standard this quantity, I chose to do so using state-level population sizes. The reason I chose population size compared to another quantity is because the entire population is at risk of being involved in a motor vehicle accident resulting in a fatality-- regardless of age or whether or not they are a driver/occupant of the motor vehicle involved.  (See the Data Section of the Supplemental Documentions )

Seasonality showed to be a non-linear variable, therefore I evaluate this variable using a natural spline to account for this.

##IV. Results

##V. Discussion


Due to that fact that motor vehicle related fatalities are 'preventable' deaths, it is a dire public health concern that various strategies at the individual, community, state, and federal level be executed to reduce these fatalities. 

##VI. Limitations
###A. Data
My data is limited. While I attempted to control for as many confounders as possible, I most definitely missed some.  

###B. Method
My chosen method involved only analyzing data from the most recent year. In this case, I cannot make any type of inference across years. I can only use my results to make statements about 2015. The reason is that while generally surrounding years are similar, trends do not always adhere to the same yearly patterns.  Essentially making inferences would indirectly suggest that I assume that all years behave the same in regards to traffic fatalities.  

##VII. Future Work
Future work that I think is vital in truly evaluating whether state traffic safety laws are associated to the amount of fatal motor vehicle accidents that occur within that state, I think it is essential to evaluate these variables throughout time.  Thus, I would consider using some type of mixed effects model to adhere to the longitudinal nature of the data. Previous research in traffic fatalities suggest that FGLS would be a good model to consider evaluating this research question.  

Pseudo experiment so that I could potentially make casual inference.  

##VIII. References & Statistical Packages

[1] Center for Disease Control and Prevention. National Center for Health Statistics. Leading Cause of Death. http://www.cdc.gov/nchs/fastats/leading-causes-of-death.htm (2016, October)

[2] Center for Disease Control and Prevention. National Center for Health Statistics. Key Injury and Violence Data.
http://www.cdc.gov/injury/wisqars/overview/key_data.html (2016, September)

[3] Blincoe, L. J., Miller, T. R., Zaloshnja, E., & Lawrence, B. A. The Economic and Societal Impact Of Motor Vehicle Crashes, 2010
Washington, DC: National Highway Traffic Safety Administration. (2015, May) 

[4] National Highway Traffic Safety Administration. Who We Are and What We Do.
http://www.nhtsa.gov/About-NHTSA/Who-We-Are-and-What-We-Do (2016, January)

[5] Cronin., Shea. The Effectiveness of Traffic Safety Laws in Reducing Fatal Traffic Accidents: A Reevaluation of State Panal Data. (p.34)
American University. Department of Justice, Law, and Society.  

[6] National Highway Traffic Safety Administration. Fatality Analysis Reporting System (FARS). Analytical User’s Manual 1975-2015

[7]

[8]

[9]

[10]

